\documentclass{report}
\usepackage{M.1.0.0}
\usepackage{hyperref}
\usepackage[spanish]{babel}
\usepackage[latin1]{inputenc}
\usepackage{multirow}
\usepackage{longtable}
\usepackage{amssymb}
\usepackage{enumitem}
%  % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % %
%
%                                                                 INICIO DEL DOCUMENTO 
% % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % %

%  % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % %
\begin{document}

\Ini

%  % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % %
%
%                                                                 Prefacio
% % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % %
\chapter*{Prefacio}

%  % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % %
%
%                                                                Indices 
% % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % %

%\tableofcontents 

%\Figura

%\Cuadros

%  % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % %
%
%                                                                Resumen 
% % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % %

%\Res{Resumen}
                  


%  % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % %
%
%                                                                Introduccion 
% % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % %
%\input{introduccion.tex} 

%  % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % %
%
%                                                               Objetivos
% % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % %
\input{objetivos.tex} 

%  % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % %
%
%                                                               Justificacion
% % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % %


\input{justificacion.tex} 
\input{problema.tex}

%  % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % %
%
%                                                               Marco Teorico
% % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % %
\input{marco-teorico.tex} 

%  % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % %
%
%                                                               Antecedentes
% % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % %
\input{antecedentes.tex} 

%  % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % %
%
%                                                               Metodologia
% % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % %
\input{metodologia.tex}

%  % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % %
%
%                                                               Resultados
% % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % %
%\input{resultados.tex}

\input{enfoqueyalcance.tex}
%  % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % %
%
%                                                       Enfoque y alcance
% % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % %

\input{diseno.tex}
%  % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % %
%
%                                                       Diseño
% % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % %

%  % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % %
%
%                                                               Analisis de Resultados
% % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % %
%\input{analisis-resultados.tex}

%  % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % %
%
%                                                              Conclusiones
% % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % %
%\input{conclusiones.tex}

%  % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % %
%
%                                                               Recomendaciones
% % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % %
%\input{recomendaciones.tex}

%  % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % %
%
%                                                               Bibliografia
% % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % %

\Biblio
\begin{thebibliography}{9}
\Bibliotoc


\bibitem[Acevedo \& Alvarado (2008)]{Acevedo2008} Acevedo, M. \& Alvarado, C. (2008). \textit{Lecciones de Semiología}, Sexta Edición. Guatemala.

\bibitem[Ambrose \& Harris (2010)]{dt-ambrose}
Ambrose, G. \& Harris, P. 2010. \textit{Design Th!nking}. Ava Publishing.

\bibitem[Becerril-Montekio \& López-Dávila (2011)]{becerril-lopez} Becerril-Montekio, Víctor y L. López-Dávila. 2011. «Sistema de Salud en Guatemala» \textit{Salud Pública Mex} \underline{LIII} (2): S197-S208.

\bibitem[California Research Bureau (2012)]{open-source-2}
California Research Bureau. 2012. \textit{Open-Source Software: Value, Cost, and Supporting Open Government} Recuperado de \url{https://www.library.ca.gov/crb/12/S-12-02.pdf}

\bibitem[Carnicero \& Fernández (2011)]{dt-carnfern} Carnicero, J. \& Fernández, A. 2011. \textit{Manual de salud electrónica para directivos de servicios y sistemas de salud}. Santiago de Chile, Naciones Unidas. 414 p\'ags.

\bibitem[Castro \& Gámez (2002)]{Castro2002}Castro, L. \& Gámez, M. (2002). \textit{Historia clínica}. In Farmacia Hospitalaria - Tomo I (pp. 295 - 305). España: Sociedad Española de Farmacia Hospitalaria. Recuperado de \url{http://www.sefh.es/bibliotecavirtual/fhtomo1/cap22.pdf}

\bibitem[Coiera (2006)]{Coiera2006}Coiera, E. (2006). \textit{Communication systems in healthcare}. The Clinical Biochemist. Reviews / Australian Association of Clinical Biochemists, 27(2), 89?98.

\bibitem[Blakely \& Timmons (2008)]{bla-timm}
Blakely, M., \& Timmons, S. 2008. \textit{Life Style and Health Research}. New York: Nova.

\bibitem[Cottom (2004)]{cottom}Cottom, Hugo. 2004. \textit{Análisis crítico del sistema nacional de salud en Guatemala}. Tesis Universidad Rafael Landívar. Quetzaltenango, Guatemala. 12 págs.

\bibitem[Dhanshetti (2015)]{dhanshetti}
Dhanshetti, S. 2015. \textit{Communication system}. Maharashtra: Industrial Training Institute.

\bibitem[Dunlop et al. (2013)]{dunlop}
Dunlop, J., Girma, D., \& Irvine, J. 2013. \textit{Digital Mobile Communications and the TETRA System}. West Sussex: John Wiley \& Sons.

\bibitem [Dutoit (1997)]{dutoit}
Dutoit, T. 1997. \textit{An Introduction to Text-to-Speech Synthesis}. (Vol. 3). Dordrecht: Springer Netherlands. http://doi.org/10.1007/978-94-011-5730-8

\bibitem [de la Harpe, Kabaso, \& Debrah (2014)]{harpe-ko}
De la Harpe, R., Kabaso, B., \& Debrah, R. 2014. \textit{Design of mobile appointment reminder and counselling system}. In Proceedings of the 9th Health Informatics in Africa Conference HELINA'14 (pp. 39-45). Koegni eHealth. Retrieved from \url{http://helina-online.org/}

\bibitem [Downer, Meara, Da Costa, \& Sethuraman (2006)]{downer-mea}
Downer, S., Meara, J., Da Costa, A., \& Sethuraman, K. 2006. \textit{SMS text messaging improves outpatient attendance}. Australian Health Review: A Publication of the Australian Hospital Association, 30(3), 389-396. \url{http://doi.org/10.1071/AH060389}

\bibitem [Egan (1993)]{egan-t}
Egan, T. 1993. \textit{Using Interactive Voice Response Systems. In Beyond Computing and Connectivity}. West Chester: DIANE Publishing Company. Pp. 391-399.

\bibitem [Emol (2015)]{bench-apple}
Emol. 2015. \textit{Apple expande tecnología para la salud en principales hospitales de Estados Unidos}. Recuperado de \url{http://www.emol.com/noticias/tecnologia/2015/02/05/702418/apple-expande-tecnologia-de-salud-en-principales-hospitales-de-estados-unidos.html}

\bibitem [Frehner (2008)]{frehner-c}
Frehner, C. 2008. \textit{Email, SMS, MMS: The Linguistic Creativity of Asynchronous Discourse in the New Media Age}. Bern: Peter Lang.

\bibitem [FrontlineSMS (2009)]{front-2009}
FrontlineSMS. 2009. \textit{Celebrating the art of the possible}. Retrieved October 1, 2015, from \url{http://www.frontlinesms.com/2009/11/19/celebrating-the-art-of-the-possible/}

\bibitem [FrontlineSMS (2011)]{front-2011}
FrontlineSMS. 2011. \textit{User Guide: Data Integrity}, 1?65. Obtenido de \url{http://www.frontlinesms.com/wp-content/uploads/2011/08/frontlinesms_userguide.pdf}

\bibitem [FrontlineSMS (2015)]{front-2015}
FrontlineSMS. 2015. \textit{FrontlineSMS Overview}. Recuperado el August 1, 2015, from \url{http://www.frontlinesms.com/technologies/frontlinesms-overview/}

\bibitem [Grover, Stewart, Lubensky, Hare, \& Jai (2012)]{grover-stewart}
Grover, A. S., Stewart, O., Lubensky, D., Hare, K., \& Jai, P. 2009. \textit{Designing Interactive Voice Response (IVR) Interfaces: localisation for low literacy users}. In Proceedings of Computers and Advanced Technology in Education (p. 8). St Thomas.

\bibitem [Gestwicki y McNely (2012)]{dt-gestwicki}
Gestwick, P. \& Mcely, B.  2013. \textit{A case study of a five-step design thinking process in educational museum game design}. Recuperado de \url{http://meaningfulplay.msu.edu/proceedings2012/mp2012_submission_37.pdf}

\bibitem [Haji, Suleman, \& Rivett (2015)]{haji-su}
Haji, H., Suleman, H., \& Rivett, U. 2015. \textit{Development of a Mobile Image-Based Reminder Application to Support Tuberculosis Treatment in Africa}. International Journal of Medical, Health, Biomedical, Bioengineering and Pharmaceutical Engineering, 9(8), 522-529.

\bibitem [Herrera (2012)]{herrera-w}
Herrera, J. 2012. \textit{Nuevas tendencias en comunicación}. 2da ed. Madrid: ESIC.

\bibitem [Hospital Roosevelt (2015)]{ho-roo}
Hospital Roosevelt. 2015. \textit{Hospital Roosevelt Gobierno de Guatemala}. Recuperado el 16 de septiembre de 2015, de \url{http://www.hospitalroosevelt.gob.gt/hr/}

\bibitem [Hospital San Juan de Dios (2015)]{ho-sanjuan}
Hospital San Juan de Dios. 2015. \textit{Hospital General San Juan de Dios}. Recuperado el 16 de septiembre de 2015, de \url{http://www.hospitalsanjuandediosguatemala.com/pages/inicio.php#.Vfnh0_l_Oko} 

\bibitem[IDEO's Attributions (2012)]{dt-ideos}
IDEO's Attributions. 2012. \textit{Design Thinking for Educators}, segunda ed. IDEO.

\bibitem [ISSSTE (2005)]{bench-issste}
ISSSTE. 2005. \textit{Sistema de cita médica telefónica e internet}. Recuperado de \url{http://189.254.143.89/issste/comun/home.aspx}

\bibitem [Huckvale, Prieto, Tilney, Benghozi, \& Car (2015)]{jose-tomas}
Huckvale, K., Prieto, J. T., Tilney, M., Benghozi, P.-J., \& Car, J. 2015. \textit{Unaddressed privacy risks in accredited health and wellness apps: a cross-sectional systematic assessment}. BMC Medicine, 13(1), 214. \url{http://doi.org/10.1186/s12916-015-0444-y}

\bibitem [Kuo, Lee, \& Tian (2006)]{kuo-lee}
Kuo, S., Lee, B., \& Tian, W. 2006. \textit{Real-Time Digital Signal Processing: Implementations and Applications}. 2nd ed., West Sussex: John Wiley \& Sons. 

\bibitem[Management Study Guide (2013)]{ManagementStudyGuide2013}
Management Study Guide. 2013. \textit{Components of Communication Process}. De \url{http://www.managementstudyguide.com/components-of-communication-process.htm}

\bibitem [Martinez-Fernandez et al. (2015)]{marti-fer} 
Martinez-Fernandez, A., Lobos-Medina, I., Diaz-Molina, C., Chen-Cruz, M., \& Prieto-Egido, I. 2015. \textit{TulaSalud: An m-health system for maternal and infant mortality reduction in Guatemala}. Journal of Telemedicine and Telecare, 21(5), 283-291. \url{http://doi.org/10.1177/1357633X15575830}

\bibitem [McDonald (2014)]{mc-donald}
McDonald, L. W. 2014. \textit{Sexual Exploitation Outreach with Text Messaging: Introducing Project Backpage}. Retrieved September 6, 2015, from \url{http://www.frontlinesms.com/2014/01/29/sexual-exploitation-outreach-with-text-messaging-introducing-project-backpage/}


\bibitem [Ministerio de Salud Pública y Asistencia Social (2013)]{hospital}
Ministerio de Salud Pública y Asistencia Social. 2013. \textit{Auditoría Financiera: Hospital Infantil de Infectología y Rehabilitación}. Recuperado el 07 de julio de 2015, de \url{http://www.mspas.gob.gt/libreacceso/images/stories/datos/2013/ABRIL%20UIP%202013/Art.%2010%20numeral%2023.%20Auditor%C3%ADas%20realizadas/CUA-22622%20(Hospital%20Infantil%20e%20Infectolog%C3%ADa).pdf}

\bibitem [Moller \& Bort (1993)]{moeller-bort}
Moeller, M. \& Bort, J. 1993. \textit{VM gets the message. Communications International}, 20, 14-15. 

\bibitem[Nist (2014)]{Nist2014}
Nist. (2014). \textit{Framework for Improving Critical Infrastructure Cybersecurity}, 39. Recuperado de \url{http://www.nist.gov/cyberframework/upload/cybersecurity-framework-021214.pdf}

\bibitem [Noar \& Harrington (2012)]{noar-harring}
Noar, S., \& Harrington, N. 2012. \textit{eHealth Applications Promising Strategies for Behavior Change}. New York: Routledge. 

\bibitem [OpenMRS (2015)]{opem-mrs-2015}
OpenMRS. 2015. \textit{Frequently Asked Questions}. Retrieved October 10, 2015, from \url{http://openmrs.org/about/faq/} 

\bibitem [Piette. et al. (2013)]{piette-mar}
Piette, J., Marinec, N., Gallegos-Cabriales, E., Gutierrez-Valverde, J., Rodriguez-Saldana, J., Mendoz-Alevares, M., \&  Silveira, M. 2013. \textit{Spanish-speaking patient?s engagement in interactive voice response (IVR) support calls for chronic disease self-management: data from three countries}. Journal of Telemedicine and Telecare, 19(2), 89-94. \url{http://doi.org/10.1177/1357633X13476234}

\bibitem [Pitkin (2014)]{pit-a}
Pitkin, A. 2014. \textit{Technology Volunteerism for Social Good; How we made Frontline easier to use}. Retrieved September 5, 2015, from \url{http://www.frontlinesms.com/2014/12/23/sc4g-multiselector/}

\bibitem [Razzouk y Shute (2012)]{dt-razzouk}
Razzouk, R. \& Shute, V.  2013. \textit{What Is Design Thinking and Why Is It Important?}. Recuperado de \url{http://myweb.fsu.edu/vshute/pdf/designthinking.pdf}

\bibitem [Salus (2013)]{bench-salus}
Salus. 2013. \textit{Salus: Software para hospitales y clínicas}. Recuperado de \url{http://www.softwaresalus.com/DefaultSalus.aspx}

\bibitem[Santizo (2010)]{santizo-firma}
Santizo, J. 2010. \textit{Implementación y adopción de la firma electrónica en Guatemala}. Universidad de San Carlos de Guatemala.

\bibitem[Scholl et al (2008)]{Scholl2008}
Scholl, M. a, Stine, K. M., Hash, J., Bowen, P., Johnson, L. A., Smith, C. D., \& Steinberg, D. I. (2008). \textit{An Introductory Resource Guide for Implementing the Health Insurance Portability and Accountability Act (HIPAA) Security Rule}, (October). Recuperado de \url{http://csrc.nist.gov/publications/nistpubs/800-66-Rev1/SP-800-66-Revision1.pdf}

\bibitem[Stair \& Reynolds (2015)]{stair-rey}
Stair, R., \& Reynolds, G. 2015. \textit{Principles of Information Systems}. 12va ed. Boston: Cengage Learning.

\bibitem [Jo Woc (2005)]{jo-woc}
Stephen Jo Woc. 2005. \textit{Ampliación Y Remodelación de la Consulta Externa de Adultos Del Hospital Roosevelt}. Universidad de San Carlos de Guatemala. Recuperado el 16 de septiembre de 2015, de \url{http://biblioteca.usac.edu.gt/tesis/02/02_1337.pdf} 

\bibitem [Stickel (2013)]{stick-c}
Stickel, C. 2013. \textit{FrontlineSMS Named 1 Tech NGO in the World by the Global Journal}. Retrieved September 1, 2015, from \url{http://www.frontlinesms.com/2013/07/24/frontlinesms-named-1-tech-ngo-in-the-world-by-the-global-journal/}

\bibitem [Swiss Tropical Institute (2015)]{swiss-trop}
Swiss Tropical Institute. 2015. eHealth Toolbox. Retrieved September 15, 2015, from http://www.swisstph.ch/services/ehealth/ehealth-toolbox.html

\bibitem [TechTimes (2015)]{bench-kit}
TechTimes. 2015. \textit{Apple HealthKit App launching in top US hospitals}. Recuperado de \url{http://www.techtimes.com/articles/31049/20150205/apple-healthkit-app-launching-in-top-us-hospitals.htm}

\bibitem [Trisby, Holley, \& Harris (2010)]{tris-h}
Trisby, F., Holley, K., \& Harris, I. 2010. \textit{Short Message Service(SMS) the creation of personal global text messaging}. Vasa. West Sussex: John Wiley \& Sons. Recuperado de \url{http://medcontent.metapress.com/index/A65RM03P4874243N.pdf}

\bibitem [Universidad de San Carlos de Guatemala (2014)]{bench-balanza}
Universidad de San Carlos de Guatemala. 2014. \textit{Evaluación de un medidor electrónico antropométrico básico para niños mayores a 6 años que automatice la lectura y registro de datos}. Recuperado de \url{http://digi.usac.edu.gt/bvirtual/informes/puicb/INF-2013-21.pdf}

\bibitem[Verizon (2015)]{Verizon2015}
Verizon. (2015). \textit{2015 Data Breach Investigations Report}. Information Security, 1-70. Recuperado de \url{http://www.verizonenterprise.com/DBIR/2015/}

\bibitem [Virji et al. (2006)]{virji-bbc}
Virji, A., Yarnall, K., Krause, K., Pollak, K., Scannell, M., Gradison, M., \& Ostbye, T. 2006. \textit{Use of email in a family practice setting: opportunities and challenges in patient- and physician-initiated communication}. BMC Medicine, 4(1), 18. \url{http://doi.org/10.1186/1741-7015-4-18}

\bibitem [Visionstate (2013)]{bench-wanda}
Visionstate. 2013. \textit{WANDA restroom management}. Recuperado de \url{http://visionstate.com/wanda/}

\bibitem [Weerawarana y Weeratunga (2004)]{opensource}
Weerawarana, S., \& Weeratunga, J. 2004. \textit{Open Source in Developing Countries}. Recuperado de \url{http://www.eldis.org/fulltext/opensource.pdf}

\bibitem [Weldon et al.(2014)]{weldon-fran}
Weldon, C., Francois, T., Trosman, J., Roggenkamp, B., Dupuy, D., Knight, J., ? Murphy, A. 2014. \textit{Abstract A84: Do patient follow-up improvements, at hospitals caring for medically underserved patients, impact no-show rates}. Cancer Epidemiology Biomarkers \& Prevention, 23(11 Supplement), A84-A84. \url{http://doi.org/10.1158/1538-7755.DISP13-A84}

\bibitem [Witten \& Madams (1977)]{witten-madams}
Witten, I. H., \& Madams, P. H. C. 1977. \textit{The telephone enquiry service: a man-machine system using synthetic speech}. International Journal of Man-Machine Studies, 9, 449-64.

\bibitem [Witter, Steele, Mcewen, \& Mehler (2001)]{witter-st}
Witter, J., Steele, A., Mcewen, D., \& Mehler, P. 2001. \textit{The Effect of Computer Generated Appointment Reminders On Compliance With Clinic Appointments}. The Internet Journal of Healthcare Administration, 8(2), 1-5. Retrieved from \url{http://ispub.com/IJHCA/8/2/10328#}

\bibitem [Wolfe \& Korytkowsk (2013)]{wolfe-kory13}
Wolfe, B., \& Korytkowsk, R. 2013.  \textit{REST Web Services Technical Documentation}. Retrieved October 4, 2015, from \url{https://wiki.openmrs.org/display/docs/REST+Web+Services+Technical+Documentation}

\bibitem [Wolfe \& Korytkowsk (2015)]{wolfe-kory}
Wolfe, B., \& Korytkowsk, R. 2015. \textit{REST Module}. Retrieved October 4, 2015, from \url{https://wiki.openmrs.org/display/docs/REST+Module}

\bibitem [Wolfe \& Purkayastha (2015)]{wolfe-purka}
Wolfe, B., \& Purkayastha, S. 2015. \textit{REST Web Services API For Clients}. Retrieved October 4, 2015, from \url{https://wiki.openmrs.org/display/docs/REST+Web+Services+API+For+Clients}

\bibitem [Yarberry (2002)]{yarberry-w}
Yarberry, W. 2002. \textit{Interactive Voice Response. In Computer Telephony Integration}. 2nd ed., pp. 79-89. Boca Ratón: CRC Press.

\bibitem [Yetisen et al. (2014)]{yetisen-m}
Yetisen, A. K., Martinez-Hurtado, J. L., da Cruz Vasconcellos, F., Simsekler, M. C. E., Akram, M. S., \& Lowe, C. R. 2014.  \textit{The regulation of mobile medical applications}. Lab on a Chip, 14(5), 833. \url{http://doi.org/10.1039/c3lc51235e}

\bibitem [6inf (2013)]{bench-his}
6inf. 2013. \textit{TESIS HIS}. Recuperado el 07 de julio de 2015, de \url{http://www.sisinf.com/es/his-software-para-hospitales.php}


\bibitem [Instituto Nacional de Estadística (2012)]{ine}
21. Instituto Nacional de Estadística. 2012. \textit{Proyecciones de INE al 30 de julio del 2012.} Recuperado de \url{http://www.ine.gob.gt/sistema/uploads/2014/02/26/5eTCcFlHErnaNVeUmm3iabXHaKgXtw0C.pdf}

\bibitem [Stead y Lin (2009)]{prefacio-2} Stead, W. y H. Lin. 2009. \textit{Computational technology for effective health care: immediate steps and strategic directions}. The National Academies Press. Recuperado de \url{https://www.nlm.nih.gov/pubs/reports/comptech_2009.pdf}

\bibitem [Fett (2000)]{techonolog-care}
Fett, M. 2000. \textit{International Technology , approaches to funding health care Care Health and Health}. Health Financing Series, 5, 42.

\bibitem [Stead \& Lin (2009)]{fett-lin}
Stead, W., \& Lin, H. S. 2009. \textit{Computational technology for effective health care immediate steps and Strategic directions}.

\end{thebibliography}


%  % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % %
%
%                                                               Anexos
% % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % %
\input{anexos.tex}

\end{document}

